A re-examination of the role of friction in the original Social Force Model

### 2019-05-22 paper.tex
barrido de todo, falta revisar los puntos que quedaron pendientes

The SFM, in its basic version, was reported to be suitable for describing, 
at least qualitatively, a variety of emergency situations, including the presence
of obstacles, or the existence of more than one exit \cite{Dorso3,Dorso5}.
More sophisticated scenarios required, however, a step up implementation
\cite{Dorso4,Dorso6,Cornes1}, although sustaining the basic model and its parameters.  \\

@@

Researchers turned back to examine the available data on the velocity and flux 
behavior for different density environments \cite{Boltes,helbing3,seyfried,seyfried1}. 
Ref.~\cite{helbing3} \textcolor{red}{summarizes} empirical data from 
literature, and their own data set, acquired from videos of the Muslim 
pilgrimage in Mina-Makkah (2006). They showed from the empirical fundamental 
diagram (flux $J$ versus density $\rho$) that highly dense crowds (seemingly up 
to $10\,$p/m$^2$) do not drive the pedestrians velocity to zero, although the 
reasons for this remain rather obscure.   \\

@@

The desired velocity for each pedestrian $i$ was 
$\vec{v}_d^{~(i)}=1$~m/s~$\hat{e}_d^{~(i)}$, where the target $\hat{e}_d^{~(i)}$ 
was set as $\hat{e}_d^{~(i)}=(L,y_i)\left \| (L,y_i) \right \|^{-1}$, being $L$ 
the \textit{x}-location of the end at the corridor and $y_i$ the 
\textit{y}-location corresponding to the \textit{ith} pedestrian 
(\textcolor{red}{see Fig.~\ref{corridor}}). This allowed the pedestrians to 
move from left to right in an unidirectional flow. Pedestrians that surpassed 
$x=L$ were re-injected at $x=0$, preserving their current velocity and 
\textit{y}-location (\textit{i.e.} periodic boundary conditions). This mechanism 
was carried out in order to keep
the crowd size unchanged.\\

@@

\textcolor{red}{The simulations were implemented on LAMMPS molecular dynamics 
simulator with parallel computing capabilities \cite{plimpton}. The time 
integration algorithm followed the velocity Verlet scheme with a time step of 
$10^{-4}$~s. All the necessary parameters were set to the same values as in 
previous works (see Refs. \cite{sticco,Dorso5}), except for the friction 
coefficient $\kappa$.} \\

@@

Researchers propose a ``re-calibration" of the (basic) SFM, in order to
attain ``stop-and-go" flows for highly dense crowds
\cite{yu1,johansson}. Presumably, this kind of instabilities within the crowd prevent
people from stopping at extremely high densities. The intended
``re-calibration" consist on either enhancing the (local) social
interactions or increasing the net-time headway (roughly, the
relaxation time) for the high density regime. Both extensions,
however, may not exclude other possibilities involving not only
individual motion, but collective (mass) motion \cite{helbing3}. Researchers
further point out that the relevance of physical contact in extremely
dense crowds may suppose a somewhat commonality with granular
media exists \cite{helbing3}.\\

@@

The former ``re-calibrations" accomplish the socio-psychological
response of the crowd to ``gain more space" (by either enhancing local
interactions or performing a delayed reaction). We are aware that
crowds may respond differently in many situations (see Ref.~\cite{drury1}). Our working
hypothesis, however, does not focus on the crowd socio-psychological
response, but on the physical contact among pedestrians (and the
walls). The socio-psychological attitude of the pedestrians will be
assumed to remain fixed along the simulations (with the 
\textcolor{red}{desired speed} limited to $v_d=1\,$m/s). \\

@@

In the case of narrow corridors, both the simulated case and the empirical results converge to a constant flow value. It is remarkable that the system does not reach a freezing state such as the one reported in Refs.\cite{lin,kwak}. Recall that our simulations do not include any respect factor (see Ref.\cite{parisi2}), or changes in the net-time headway (Ref.\cite{helbing3}), or the urge to see an attraction (Ref.\cite{kwak}). We assume a well defined target and the same $v_d$ for all the pedestrians.\\

@@

As our desired velocity is fixed at $v_d = 1$~m/s, we conclude that the speed at 
the maximum density has to be bounded by $v(\rho_{max}) \lesssim  0.5$~m/s in 
order to satisfy the qualitative behavior of the (experimental) fundamental 
diagram reported in the literature.\\

The above reasoning is consistent with the speed-density results shown in Fig.~\ref{fundamental_diagram_speed}. As a visual guide we plotted $v=0.5$~m/s with a horizontal dashed line. The close examination of $\rho_{max} = 9$ shows that values corresponding to the wide corridors ($w=15$~m and $w=22$~m) exceed $v=0.5$~m/s. But, those values corresponding to narrow corridors fall below $v=0.5$~m/s. \\

@@

The results shown in Fig.~\ref{fundamental_diagram_speed} confirm the fact that when the density is low enough, pedestrians manage to walk at the desired velocity ($v=v_d=1$~m/s). Above $\rho>5$, however, the velocity begins to slow down. The inset shows the experimental data at the entrance of the Jamaraat bridge. We may conclude that our simulations agree with the experimental data for narrow corridors, but disagree as these become wider. The wider the corridor, the greater the velocity for all the density values explored. In Section \ref{velocity_profile} we will further discuss this topic.\\

@@

As we mentioned in Section \ref{fundamental_diagram}, when the density is low, pedestrians achieve the desired velocity ($v=v_d=1$~m/s). Since the results of the previous Section only hold for the area located in the middle of the corridor (see \textcolor{red}{dashed circle in} Fig.~\ref{corridor}), we want to shed some light and understand what is happening across the entire corridor.\\

@@

In the three figures we observe a similar pattern: the regions near the walls (bottom and top) are the most dissipative ones. The center of the corridor is though not a very dissipative region. Furthermore, the work seems to increase with the corridor width. This occurs because the relative velocity between pedestrians is greater in the wide corridors than in narrow ones. This can be checked by comparing the slopes between parabolas in Fig.~\ref{speed-profile-multi_width}. As can be seen there, the wider the corridor, the greater the slope (corresponding to locations near the wall).\\

@@

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.5\columnwidth]{./abswg_width10.eps}\label{abswg_width10}}\\ 
    \subfloat[]{\includegraphics[width=0.5\columnwidth]{./abswg_width15.eps}\label{abswg_width15}}\\
    \subfloat[]{\includegraphics[width=0.5\columnwidth]{./abswg_width22.eps}\label{abswg_width22}}\
\caption[width=0.47\columnwidth] {Color map for the absolute value of the work done by the friction. The density chose was $\rho=6$. (a) Corresponds to $w=$10~m (b) $w=$15~m and (c) $w=$22~m. The axis represent the location in the corridor ($x$ and $y$). The scale bar on the right is expressed in Joule units. The work was numerically integrated following the Trapezoidal rule with $\Delta t =0.05$~s. The pedestrians desired velocity was $v_d = 1$~m/s. The contour lines were computed on a square grid of 1m$\times$~1m and then splined to get smoother curves.} 
\label{abswg}
\end{figure}

@@

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\columnwidth]
{./fracc_clusteriz_vs_density.eps}
\caption{\label{fic} Fraction of clustered individuals as a function of the density. Squared symbols correspond to $\kappa_w=\kappa_i=2.4\times 10^6$, circles to $\kappa_i=2.4\times 10^6$ and $\kappa_w=2.4\times 10^5$, triangles to to $\kappa_i=2.4\times 10^5$ and $\kappa_w=2.4\times 10^6$ and crosses to the coefficients of the original Social Force Model ($\kappa_i=2.4\times 10^5$ and $\kappa_w=2.4\times 10^5$). The pedestrians walk across a corridor with $v_d=1$m/s. The measurements were recorded every 0.5~s once the system reached the stationary state. The values corresponding to the fraction of clustered individuals were averaged over 170~s. The cluster cutoff distance was 0.46~m (equivalent to the shoulder width of the pedestrians).}
\end{figure}

@@

@@

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\columnwidth]
{./fis_kx10.eps}
\caption{\label{fis} Mean evacuation time (seconds) vs. the pedestrian’s desired velocity (m/s). The room was 20 m $\times$ 20 m size. Mean values were computed from 30 evacuation processes. The pedestrians were initially placed in a regular square arrangement along the room with random velocities, resembling a Gaussian distribution with null mean value. The friction coefficient was $\kappa=2.4\times10^{6}$. The inset shows the evacuation time vs the desired velocity for the friction coefficient corresponding to the original model ($\kappa=2.4\times10^{5}$).  }
\end{figure}

>>>>

The SFM, in its basic version, was reported to be suitable for describing, 
at least qualitatively, a variety of emergency situations, including the presence
of obstacles, or the existence of more than one exit \cite{Dorso3,Dorso5}.
More sophisticated scenarios required, however, a step up implementation
\cite{Cornes1,Dorso4,Dorso6}, although sustaining the basic model and its parameters.  \\

@@

Researchers turned back to examine the available data on the velocity and flux 
behavior for different density environments \cite{Boltes,helbing3,seyfried1,seyfried}. 
Ref.~\cite{helbing3} \textcolor{red}{summarizes} empirical data from 
literature, and their own data set, acquired from videos of the Muslim 
pilgrimage in Mina-Makkah (2006). They showed from the empirical fundamental 
diagram (flux $J$ versus density $\rho$) that highly dense crowds (seemingly up 
to $10\,$p~m$^{-2}$) do not drive the pedestrians velocity to zero, although the 
reasons for this remain rather obscure.   \\

@@

The desired velocity for each pedestrian $i$ was 
$\vec{v}_d^{~(i)}=1$~m~s$^{-1}$~$\hat{e}_d^{~(i)}$, where the target $\hat{e}_d^{~(i)}$ 
was set as $\hat{e}_d^{~(i)}=(L,y_i)\left \| (L,y_i) \right \|^{-1}$, being $L$ 
the \textit{x}-location of the end at the corridor and $y_i$ the 
\textit{y}-location corresponding to the \textit{ith} pedestrian 
(\textcolor{red}{see Fig.~\ref{corridor}}). This allowed the pedestrians to 
move from left to right in an unidirectional flow. Pedestrians that surpassed 
$x=L$ were re-injected at $x=0$, preserving their current velocity and 
\textit{y}-location (\textit{i.e.} periodic boundary conditions). This mechanism 
was carried out in order to keep
the crowd size unchanged.\\

@@

\textcolor{red}{The simulations were implemented on LAMMPS molecular dynamics 
simulator with parallel computing capabilities \cite{plimpton}. The time 
integration algorithm followed the velocity Verlet scheme with a time step of 
$10^{-4}$~s. All the necessary parameters were set to the same values as in 
previous works (see Refs. \cite{Dorso5,sticco}), except for the friction 
coefficient $\kappa$.} \\

@@

Researchers propose a ``re-calibration" of the (basic) SFM, in order to
attain ``stop-and-go" flows for highly dense crowds
\cite{johansson,yu1}. Presumably, this kind of instabilities within the crowd prevent
people from stopping at extremely high densities. The intended
``re-calibration" consist on either enhancing the (local) social
interactions or increasing the net-time headway (roughly, the
relaxation time) for the high density regime. Both extensions,
however, may not exclude other possibilities involving not only
individual motion, but collective (mass) motion \cite{helbing3}. Researchers
further point out that the relevance of physical contact in extremely
dense crowds may suppose a somewhat commonality with granular
media exists \cite{helbing3}.\\

@@

The former ``re-calibrations" accomplish the socio-psychological
response of the crowd to ``gain more space" (by either enhancing local
interactions or performing a delayed reaction). We are aware that
crowds may respond differently in many situations (see Ref.~\cite{drury1}). Our working
hypothesis, however, does not focus on the crowd socio-psychological
response, but on the physical contact among pedestrians (and the
walls). The socio-psychological attitude of the pedestrians will be
assumed to remain fixed along the simulations (with the 
\textcolor{red}{desired speed} limited to $v_d=1\,$ m~s$^{-1}$). \\

@@

In the case of narrow corridors, both the simulated case and the empirical results converge to a constant flow value. It is remarkable that the system does not reach a freezing state such as the one reported in Refs.\cite{kwak,lin}. Recall that our simulations do not include any respect factor (see Ref.\cite{parisi2}), or changes in the net-time headway (Ref.\cite{helbing3}), or the urge to see an attraction (Ref.\cite{kwak}). We assume a well defined target and the same $v_d$ for all the pedestrians.\\

@@

As our desired velocity is fixed at $v_d = 1$~m~s$^{-1}$, we conclude that the speed at 
the maximum density has to be bounded by $v(\rho_{max}) \lesssim  0.5$~m~s$^{-1}$ in 
order to satisfy the qualitative behavior of the (experimental) fundamental 
diagram reported in the literature.\\

The above reasoning is consistent with the speed-density results shown in Fig.~\ref{fundamental_diagram_speed}. As a visual guide we plotted $v=0.5$~m~s$^{-1}$ with a horizontal dashed line. The close examination of $\rho_{max} = 9$ shows that values corresponding to the wide corridors ($w=15$~m and $w=22$~m) exceed $v=0.5$~m~s$^{-1}$. But, those values corresponding to narrow corridors fall below $v=0.5$~m~s$^{-1}$. \\

@@

The results shown in Fig.~\ref{fundamental_diagram_speed} confirm the fact that when the density is low enough, pedestrians manage to walk at the desired velocity ($v=v_d=1$~m~s$^{-1}$). Above $\rho>5$, however, the velocity begins to slow down. The inset shows the experimental data at the entrance of the Jamaraat bridge. We may conclude that our simulations agree with the experimental data for narrow corridors, but disagree as these become wider. The wider the corridor, the greater the velocity for all the density values explored. In Section \ref{velocity_profile} we will further discuss this topic.\\

@@

As we mentioned in Section \ref{fundamental_diagram}, when the density is low, pedestrians achieve the desired velocity ($v=v_d=1$~m~s$^{-1}$). Since the results of the previous Section only hold for the area located in the middle of the corridor (see \textcolor{red}{dashed circle in} Fig.~\ref{corridor}), we want to shed some light and understand what is happening across the entire corridor.\\

@@

In the three figures we observe a similar pattern: the regions near the walls (bottom and top) are the most dissipative ones. The center of the corridor is though not a very dissipative region. Furthermore, the work seems to increase with the corridor width. This occurs because the relative velocity between pedestrians is greater in the wide corridors than in narrow ones. The wider the corridor, the greater the slope (corresponding to locations near the wall).\\

@@

\begin{figure}[!htbp]
\centering
    \subfloat[]{\includegraphics[width=0.5\columnwidth]{./abswg_width10.eps}\label{abswg_width10}}\\ 
    \subfloat[]{\includegraphics[width=0.5\columnwidth]{./abswg_width15.eps}\label{abswg_width15}}\\
    \subfloat[]{\includegraphics[width=0.5\columnwidth]{./abswg_width22.eps}\label{abswg_width22}}\
\caption[width=0.47\columnwidth] {Color map for the absolute value of the work done by the friction. The density chose was $\rho=6$. (a) Corresponds to $w=$10~m (b) $w=$15~m and (c) $w=$22~m. The axis represent the location in the corridor ($x$ and $y$). The scale bar on the right is expressed in Joule units. The work was numerically integrated following the Trapezoidal rule with $\Delta t =0.05$~s. The pedestrians desired velocity was $v_d = 1$~m~s$^{-1}$. The contour lines were computed on a square grid of 1m$\times$~1m and then splined to get smoother curves.} 
\label{abswg}
\end{figure}

@@

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\columnwidth]
{./fracc_clusteriz_vs_density.eps}
\caption{\label{fic} Fraction of clustered individuals as a function of the density. Squared symbols correspond to $\kappa_w=\kappa_i=2.4\times 10^6$, circles to $\kappa_i=2.4\times 10^6$ and $\kappa_w=2.4\times 10^5$, triangles to to $\kappa_i=2.4\times 10^5$ and $\kappa_w=2.4\times 10^6$ and crosses to the coefficients of the original Social Force Model ($\kappa_i=2.4\times 10^5$ and $\kappa_w=2.4\times 10^5$). The pedestrians walk across a corridor with $v_d=1$m~s$^{-1}$. The measurements were recorded every 0.5~s once the system reached the stationary state. The values corresponding to the fraction of clustered individuals were averaged over 170~s. The cluster cutoff distance was 0.46~m (equivalent to the shoulder width of the pedestrians).}
\end{figure}

@@

\textcolor{red}{We interpret $\mathcal{A}$ as a parameter related exclusively to the 
repulsion between pedestrians and $\mathcal{K}$  as parameter related to the friction-repulsion.
While the former is valid in every situation, the latter only takes part when the
pedestrians are in contact.}\\

@@

\begin{figure}[htbp!]
\centering
\includegraphics[width=0.7\columnwidth]
{./fis_kx10.eps}
\caption{\label{fis} Mean evacuation time (seconds) vs. the pedestrian’s desired velocity (m~s$^{-1}$). The room was 20 m $\times$ 20 m size. Mean values were computed from 30 evacuation processes. The pedestrians were initially placed in a regular square arrangement along the room with random velocities, resembling a Gaussian distribution with null mean value. The friction coefficient was $\kappa=2.4\times10^{6}$. The inset shows the evacuation time vs the desired velocity for the friction coefficient corresponding to the original model ($\kappa=2.4\times10^{5}$).  }
\end{figure}
